Epidemiological birth-death models with partner notification

### 2024-07-05 article/article_plos.tex
Version 0.1.3
Article abstract

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Epidemiological birth-death models with partner notification} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anna Zhukova\textsuperscript{1,2,3*},
Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
Olivier Gascuel\textsuperscript{1,4},
\\
\bigskip
\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{2} Bioinformatics and Biostatistics Hub, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{3} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{4} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
\\
\bigskip

% Current address notes
\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Curabitur eget porta erat. Morbi consectetur est vel gravida pretium. Suspendisse ut dui eu ante cursus gravida non sed sem. Nullam sapien tellus, commodo id velit id, eleifend volutpat quam. Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus. Nunc pellentesque pretium diam, quis maximus dolor faucibus id. Nunc convallis sodales ante, ut ullamcorper est egestas vitae. Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e. genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g. due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g. quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g. some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g. due to their immune system particularities or to their behaviour).

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BDS model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g. SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g. Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g. HIV) more than the others~\cite{Stadler2014}.

To allow for heterogeneity over time, Stadler \textit{et al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them. Kühnert \textit{et al.}~{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types. In particular the skyline approach can be useful to account for changes in the sampling policies (e.g. before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).

Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling. We can propose different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability over time intervals with the BDSKY, however, none of them can account for correlation effects in sampling, in particular due to partner notification (contact tracing). Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.

In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN. We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models. We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees. For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator. We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemic in the UK.

\section*{Materials and methods}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
while bifurcations (i.e. internal nodes) correspond to pathogen transmissions (Fig.~\ref{fig:tt}). The tree branches are measured in units of time. Let $T$ denote the time that passed between the tree root (the beginning of the (sub-)epidemic, at time $t=0$) and the last sampled tip (at $t=T$).

\begin{figure}[!h]
\centering 
%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
\includestandalone[width=0.6\textwidth]{Fig_tree}
\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e. tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g. $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
\label{fig:tt} 
\end{figure}

@@

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g. by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

@@

In this section we propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$. The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to the health officials for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g. got diagnosed though the standard procedure), they get tested upon notification. The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data. Upon notification, the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$). Upon sampling, the partner might notify in their turn. BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via standard procedure) gets removed \textit{and observed} with this rate. This creates an observed tip in the transmission tree. 
\end{itemize}

@@

\item notified individuals will eventually get sampled (i.e., their pathogens are always observed upon removal, unlike the standard case where the observation happens with a sampling probability $\rho$);\\

Realistically again, we assume here that the diagnosis and follow-up process works as expected, sampling notified patients.

@@

This assumption might need to be removed or relaxed (e.g. to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work. It is however useful to make the mathematics and likelihood calculation under the model manageable. In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g. spouse).

Note that the ``most recent partner'' relationship is not necessarily symmetrical. The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$ and before the notification by $i$. Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.  
\end{enumerate}

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in~\nameref{S1_Appendix}. The results in the following section show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

@@

\subsection*{Testing for presence of Partner Notification}~\label{sec:test}
To access whether a given tree reveals the presence of partner notification, we developed a non-parametric test based on cherries (two-tip clades).

@@

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time. Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$. Hence the external branches that start close to the end of the sampling period would be on average shorter (as if they are not sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

Let us assume that there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$. We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}. If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle (see Fig.~\ref{fig:tipswap}). We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries. 
The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g. to month or even a year). Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries. However, if some cherries are ``notified'' their real tip differences should be smaller than reshuffled ones. We therefore output the pvalue calculated by the sign test.

\begin{figure*}[tbhp]
\centering 
\includegraphics[width=0.5\textwidth]{Fig_cherryswap.png}
\caption{Reshuffled cherry generation. \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left). The randomly selected tip is shown with a bold branch for each cherry (e.g. the tip on the right for the leftmost cherry). The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$). \textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth. As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle. The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
\label{fig:tipswap} 
\end{figure*}

@@

We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator (see section~"\nameref{sec:sim}").

@@

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$, starting on this branch at a time $t \leq t_i$ till time $t_i$. The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree. $t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch). % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

@@

In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal). Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.  Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$. $p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Epidemiological birth-death models with partner notification} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Anna Zhukova\textsuperscript{1,2,3*},
%Mathieu Moslonka-Lefebvre\textsuperscript{1\textcurrency},
Olivier Gascuel\textsuperscript{1,4},
\\
\bigskip
\textbf{1} Unit\'e Bioinformatique Evolutive, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{2} Bioinformatics and Biostatistics Hub, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{3} G5 Evolutionary Dynamics of Infectious Diseases, Institut Pasteur, Université de Paris, Paris, France
\\
\textbf{4} Institut de Syst\'{e}matique, Evolution, Biodiversit\'{e} (ISYEB) - URM 7205 CNRS, Mus\'{e}um National d'Histoire Naturelle, SU, EPHE \& UA, Paris, France
\\
\bigskip

% Current address notes
%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
% \textcurrency b Insert second current address 
% \textcurrency c Insert third current address

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Phylodynamics bridges the gap between classical epidemiology and pathogen genome sequence data by estimating epidemiological parameters from time-scaled pathogen phylogenetic trees. The models used in phylodynamics typically assume that the sampling procedure is independent between infected individuals. However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification schemes are included in health policies of many countries.

We developed an extension of phylodynamic multi-type birth-death (MTBD) models with partner notification (PN), and a simulator to generate trees under MTBD and MTBD-PN models. We proposed a non-parametric test for detecting partner notification in pathogen phylogenetic trees. Its application to simulated data showed that it is both highly specific and sensitive. For the simplest representative of the MTBD-PN family, the BD-PN model, we developed a maximum likelihood parameter and CI estimator. It performed accurate estimations on simulated data, and detected partner notification in HIV-1 B epidemics in Zurich and the UK. Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

Our PN test, MTBD-PN tree simulator and BD-PN parameter estimator are freely available at \href{github.com/evolbioinfo/treesimulator}{https://github.com/evolbioinfo/treesimulator} and \href{github.com/evolbioinfo/bdpn}{https://github.com/evolbioinfo/bdpn}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Phylodynamic models can estimate epidemiological parameters from pathogen phylogenetic trees (i.e., genealogies). These models typically assume that sampling of pathogen sequences from detected cases is independent between infected individuals. However this assumption does not hold for many epidemics, in particular for such sexually transmitted infections as HIV-1, for which partner notification (contact tracing) is included in health policies of many countries.

We developed a phylodynamic model accounting for partner notification and a test for detection of partner notification in pathogen phylogenetic trees. They showed good performance both on simulated and real data (HIV-1 B epidemics in Zurich and the UK). Importantly, we showed that not accounting for partner notification when it is present leads to bias in parameter estimation, while our new estimator performs well both in presence and in absence of partner notification.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
The interaction of epidemiological and evolutionary processes leaves a footprint in pathogen genomes. Phylodynamics leverages this footprint to estimate epidemiological parameters~\cite{Grenfell2004a,Volz2013}. It relies on models that bridge the gap between traditional epidemiology and sequence data by estimating such parameters as the basic reproduction number, $R_0$, from topology and branch lengths of pathogen phylogenies (i.e., genealogies of the pathogen population, approximating the transmission trees) combined with metadata on the samples. Classical epidemiological methods make inferences from incidence curves, which in the presence of import cases might be biased and give appearance of elevated local transmission. In contrast, phylodynamic approach can overcome such biases as it uses the transmission structure as one of its information sources~\cite{vaughanEstimatesEarlyOutbreakspecific2024}. Therefore, using genomic data with phylodynamic estimations can provide valuable insights for preventing epidemic spreads.

Phylodynamic models can be classified into two main families:  coalescent~\cite{Volz2009a,Drummond2005,Pybus2000a} and birth-death (BD)~\cite{Kendall1948,Maddison2007,Stadler2009,Stadler2010}. Coalescent models are often preferred for estimating deterministic population dynamics, however for highly stochastic processes, such as the dynamics of emerging pathogens, BD models are better adapted~\cite{Macpherson2021}. In the classic BD model with incomplete sampling~\cite{Stadler2009}, births represent pathogen transmission events (happening at a constant transmission rate), while deaths correspond to becoming non-infectious (e.g., due to healing, self-isolation, starting a treatment, or death, modelled with a constant removal rate).

The epidemic spread and detection can however be non-homogenous and depend on different factors, including governmental health policies (e.g., quarantine measures or development of pathogen detection tests), pathogen evolution over time (e.g., some SARS-CoV-2 variants are more transmissible than the others),  differences between host individuals (e.g., due to their immune system particularities or to their behaviour).

To allow for heterogeneity at the population level, the multi-type birth-death (MTBD) extention of the classical BD model was developed by Stadler \textit{et al.}~\cite{Stadler2013a}. MTBD framework allows for different types of individual states (and changes between them, modelled with state change rates). These models are phylodynamic analogies of compartmental models in classical epidemiology (e.g., SIR, Susceptible-Infectious-Removed).  Examples of the MTBD family models include the birth-death exposed-infectious (BDEI) model, which was designed for pathogens featuring an incubation period between the moments of infection and of becoming infectious (e.g., Ebola and SARS-CoV-2), and the birth-death with superspreading model (BDSS) accounting for the fact that some individuals might spread the pathogen (e.g., HIV) more than the others~\cite{Stadler2014}.

To allow for heterogeneity over time, Stadler \textit{et al.}~\cite{Stadler2013} developed one-state Bayesian birth-death skyline plot (BDSKY) that divides the time into intervals and allows for different piecewise constant rates on them. Kühnert \textit{et al.}~\cite{Kuhnert2016} combined the MTBD model with the BDSKY to allow for both piecewise-constant rate changes over time and multiple individual types. In particular the skyline approach can be useful to account for changes in the sampling policies (e.g., before/after HIV discovery or before/after PCR test spread for SARS-CoV-2).

Another important source of heterogeneity that none of these models accounts for is inflicted by non-independent sampling. One can set different sampling probabilities for different types of individuals in MTBD framework, or change of the sampling probability between time intervals with the BDSKY, however, none of these models can account for correlation effects in sampling, in particular due to partner notification (contact tracing). Partner notification or contact tracing is a process in which contacts of detected infected individuals are identified and offered testing for infection. It has been applied to various infectious diseases, including syphilis, HIV, tuberculosis, Ebola and COVID-19, and shown to successfully reduce transmission~\cite{el-sadrContactTracingBarriers2022}. For instance, for HIV,  partner notification showed increased early referral and initiation of treatment, and several countries include it in their national HIV testing services policies~\cite{worldhealthorganizationCountryPolicyReview2016}.

In this study we propose an extension of MTBD models that allows for modeling non-random sampling due to partner notification (PN), MTBD-PN. We describe the -PN extension and its assumptions, and present a transmission tree simulator under MTBD-PN models. We propose a non-parametric test for detecting partner notification in pathogen phylogenetic trees. For the simplest representative of this model family, the BD-PN model, we also derive the equations for its likelihood calculation and present a maximum likelihood parameter estimator. We test our BD-PN parameter estimator on simulated data, and apply it to the study of HIV-1 B epidemics in Zurich and in the UK. Importantly, we show that not accounting for partner notification when it is present leads to bias in parameter estimation with the BD model, while BD-PN parameter estimator performs well both in presence and in absence of partner notification.

\section*{Materials and methods}
In a pathogen transmission tree $\mathscr{T}$ (approximated by a time-scaled pathogen phylogeny) the tips represent sampled pathogens %, patient state transitions occur along the branches, 
while bifurcations (i.e., internal nodes) correspond to pathogen transmissions. % (Fig.~\ref{fig:tt}). 
The tree branches are measured in units of time, and the time goes forward, from the tree root (representing the beginning of the (sub-)epidemic, at time $t=0$) to the time of the last sampled tip (at $t=T$).

%\begin{figure}[!h]
%\centering 
%%\includegraphics[width=0.8\textwidth]{Transmission_tree.png}
%\includestandalone[width=0.6\textwidth]{Fig_tree}
%\caption{\textbf{Transmission tree} $\mathscr{T}$ with $n=5$ external nodes (i.e., tips, which correspond to sampling events: $0000, 0001, 001, 010, 011$), $n-1=4$ internal nodes (which correspond to transmissions: $0$ (the root) and $00, 01, 000$) and $2n - 2 = 8$ branches (plus the root branch of zero length). %The lengths of the branches are shown on the right of each branch: $t_i$ is the lengths of the branch connecting the node $i$ to its parent. 
%Time $t$ starts at the root of the tree ($t=0$) and goes till the last sampled tip. The times of the nodes are shown on the left, e.g., $t_{0001}$ is the time of tip $0001$ (when $0001$'s pathogen was sampled). $T$ corresponds to the end of the sampling period (when the most recent tip, $0000$, was sampled).}
%\label{fig:tt} 
%\end{figure}

@@

Under this model we can calculate the likelihood density $L(\mathscr{T}|\Theta)$ of a tree $\mathscr{T}$ under parameter values $\Theta = \{\lambda, \psi, \rho\}$ (see \nameref{math}). The likelihood density can then be used to estimate model parameters from a phylogenetic tree (e.g., by finding parameter values that maximize it, in maximum likelihood framework). In particular, it permits inference of the following epidemiological parameters:

@@

We propose an %non-Markovian 
extension of the MTBD models that adds partner notification (PN).  In the simplest case, the BD-PN model, at the moment of sampling the sampled individual might notify their most recent partner with a constant probability $\upsilon$. The notification implies that four conditions are met: (i) the sampled individual remembers their last partner and accepts to either notify the partner on their own or to give the partner's contacts to health-care providers for assisted notification; (ii) a pathogen transmission happened at the moment of contact (either the sampled individual infected the partner or the other way round); (iii) the partner is successfully reached and notified; (iv) unless the partner already became uninfectious by other means by the moment of notification (e.g., got diagnosed though the standard procedure), they get tested upon notification. The parameter $\upsilon$ hence regroups a complex set of conditions, which would hardly be identifiable separately from only sequence data. Upon notification, the partner's pathogen is sampled almost instantaneously (modeled via a constant notified sampling rate $\phi >> \psi$). Upon sampling, the partner might notify in their turn. BD-PN hence has 3 classical BD and 2 additional parameters:
\begin{itemize}
 \item $\lambda$ -- constant transmission rate: an infected individual transmits their pathogen to a newly infected individual (which corresponds to an internal node in the transmission tree);
 \item $\psi$ -- constant removal rate: an infected individual stops being infectious at this rate. This corresponds to a tip in the transmission tree (observed or hidden);
 \item $\rho$ -- constant sampling probability: upon removal the individual's pathogen might get sampled with this probability, which corresponds to an observed tip in the transmission tree;
 \item $\upsilon$ -- constant notification probability: upon sampling, the sampled individual might notify their most recent partner with this probability;
 \item $\phi >> \psi$ -- constant notified sampling rate: a notified individual (if not yet removed via the standard procedure) gets removed \textit{and observed} with this rate. This creates an observed tip in the transmission tree. 
\end{itemize}

@@

\item successfully notified individuals get sampled (i.e., their pathogens are always observed upon removal, unlike the standard case where the observation happens with a sampling probability $\rho$);\\

Realistically again, we assume here that the diagnostic and follow-up process works as expected, sampling notified patients.

@@

This assumption might need to be removed or relaxed (e.g., to up to a certain fixed number of most recent partners, or all the partners within a certain time interval) in the future work. It is however useful to make the mathematics and likelihood calculation under the model manageable. In the case of sexually transmitted diseases, notifying only the most recent partner could roughly correspond to notifying the ``main'' partner (e.g., spouse).

Note that the ``most recent partner'' relationship is not necessarily symmetrical. The most recent partner $j$ of individual $i$ might have transmitted their virus further (to individual $k$) after the contact with $i$. Hence, while $j$ is the most recent partner of $i$, $j$'s most recent partner is $k$, and, once notified, $j$ might notify $k$.  
\end{enumerate}

@@

We implemented a tree simulator that generates sampled transmission trees under MTBD and MTBD-PN models (with $m$ states). It is Gillespie-based, generates state change, transmission and removal times, and only reconstructs the sampled parts of the tree to save memory and increase speed.
We give its algorithmic details in~\nameref{S1_Appendix}. The results in section~"\nameref{sec:test:sim}" show that the trees generated under MTBD and MTBD-PN models are different and easily distinguishable.

@@

\subsection*{Testing for presence of Partner Notification}~\label{sec:test}
To access whether a given tree reveals partner notification, we developed a non-parametric test based on cherries (two-tip clades).

@@

More formally, in the absence of partner notification (null hypothesis, H0), we assume that branch lengths of two cherry tips are independently drawn from the same distribution (which is the case for MTBD models). In the presence of partner notification (alternative hypothesis, H1), we assume that for some of the cherries (including notified partners) their branch lengths are not independent and are similar. Therefore, in the absence of partner notification two external branches taken randomly from the tree and two external branches belonging to the same cherry should be indistinguishable, if they all start at about the same time. Being close in starting time is important as the tree is censored by the end of the sampling period: $t \leq T$. Hence the external branches that start close to the end of the sampling period would be on average shorter (as unless sampled before $T$ they will stay hidden) than those that start closer to the root~\cite{mooersBranchLengthsBirth2012}. %We give control to the user on what should be considered ``some of the cherries'' and ``similar branch lengths'' for their data, via adjustable test parameter values.

Let us assume that there are N cherries in the tree. For each cherry $i\;(1 \leq i \leq N)$ with tips $i0$ and $i1$, we calculate the difference between its tip times $d_i = |t_{i0} - t_{i1}|$. We then generate a collection of reshuffled cherries by (1) ordering the cherries by their root dates; (2) for each cherry $i$ randomly selecting one of its tips $i* \in \{i0, i1\}$; (3) swapping the selected tips between the neighboring cherries: the tip $2i*$ gets swapped with the tip $2i-1*$ ($1 \leq i \leq N / 2$) as illustrated in Fig.~\ref{fig:tipswap}. (If the total number of cherries N is odd, instead of swapping the selected tips of the last two cherries, we swap the selected tips of the last three cherries in a cycle as in Fig.~\ref{fig:tipswap}). We then calculate the time differences $d_{i'}$ between the tips of these shuffles cherries. 
The test then compares $d_i$ to $d_{i'}$: $sign(d_{i'} - d_i) = 1$ if $d_{i'}$ is larger than $d_i$; $= -1$ if it is smaller; or $= 0$ if they are equal (this could happen as in real datasets dates are often truncated, e.g., to month or even a year). Note that if cherry tip branch lengths are independently drawn from the same distribution, $d_{i'}$ will be smaller than $d_i$ for about half of the cherries. However, if some cherries are ``notified'' their real tip differences should be smaller than reshuffled ones. We therefore output the pvalue calculated by the sign test.

\begin{figure*}[tbhp]
\centering 
\includegraphics[width=0.5\textwidth]{Fig_cherryswap.png}
\caption{Reshuffled cherry generation. \textbf{a.} Cherries of the original tree sorted by their root times (times of the roots of the two oldest cherries, $t_{r1}$ and $t_{r2}$, are shown on the left). The randomly selected tip is shown with a bold branch for each cherry (e.g., the tip on the right for the leftmost cherry). The tip sampling time differences are shown for the two leftmost cherries ($d_1$ and $d_2$). \textbf{b.} Reshuffled cherries, which were obtained from the original ones by swapping the selected tips between the neighbouring cherry couples (as shown with arrows in a.): the first and the second; the third and the forth. As the total number of cherries (7) is odd, the last three (instead of two) cherries exchanged their selected tips in a cycle. The reshuffled tip sampling time differences are shown for the two leftmost cherries ($d'_1$ and $d'_2$).}
\label{fig:tipswap} 
\end{figure*}

@@

We implemented the PN-test in Python 3. It is available as a part of BDPN parameter estimator ($pn\_test$, see section~"\nameref{sec:sim}").

@@

With the equation~(\ref{eq:Us}), we can write down the equation for the probability density $p^{(i)}(t)$ of evolving as on an observed tree branch $i$ till time $t_i$, starting on this branch at time $t \leq t_i$. The master equations for the BD models were initially developed by Stadler \textit{et al.}~\cite{Stadler2009}, however what we present here is their branch-specific formulation that we proposed in~\cite{zhukovaFastAccurateMaximumLikelihood2022}:
In our formulation, ${p}^{(i)}(t)$ describes an evolution along a tree branch $i$, without taking into account the branch's subtree. $t_i$ corresponds to the time of the equation's initial condition (typically at the end of the branch $i$). % and provided that we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at time $t$. These two individuals could be different if a hidden transmission, where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$.

@@

In the classical BD model with incomplete sampling, a branch of the tree corresponds to a virus evolution between the transmission corresponding to the node at the beginning of this branch and either a sampling event (if the branch is external) or a transmission corresponding to the node at the end of the branch (if it is internal). Note, that since sampling is incomplete, hidden transmissions (leading to fully unsampled subtrees) could occur along this branch, where the unsampled subtree could correspond to either donor or recipient.  Hence, we do not know whether the individual at the end of the branch (at time $t_i$) is the same as at any time $t < t_i$ along the branch. These two individuals could be different if a hidden transmission where the donor subtree stayed unobserved, occurred between times $t$ and $t_i$. $p^{(i)}(t)$ in equations~(\ref{eq:p}) describes an evolution along such a tree branch, integrating over all the possibilities, starting from no hidden transmission along the branch and including any number of hidden transmissions with either donor or recipient tree staying unsampled (and hence potentially different individuals at times $t$ and the branch end $t_i$).
